---
title: "Biostatistics for fluid biomarkers"
author: 
  name: Michael C Donohue PhD
  company: USC Alzheimer's Therapeutic Research Institute
date: April 20, 2020
logo: "images/usc.png"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## About me

* 2001 - 2005: PhD Mathematics, University of California, San Diego
  - *Rank Regression & Synergy Detection*
* 2005 - 2015: Biostatistician/Faculty, University of California, San Diego
  - Alzheimer's Disease Neuroimaging Initiative (ADNI)
  - Alzheimer's Disease Cooperative Study (ADCS)
* 2015 - Present: Faculty, University of Southern California, San Diego
  - Associate Director of Biostatistics, [Alzheimer's Therapeutic Research Institute (ATRI)](https://keck.usc.edu/atri/)
  - Biostatistics Unit Co-Lead, [Alzheimer's Clinical Trial Consortium (ACTC)](https://www.actcinfo.org/)

## Course Overview

Topics:

* 9:00 - 9:50 -- Biostatistics for Fluid Biomarkers
* 10:00 - 10:50 -- Biostatistics for Imaging Biomarkers
* 11:00 - 11:50 -- Modeling Longitudinal Data

Emphases:

* Visuliazation 
* Demonstrations using R, code available from:
  - [https://github.com/atrihub/biomarkers-neuro-disorders-2020](https://github.com/atrihub/biomarkers-neuro-disorders-2020)


## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
# library(nCal)
# library(calibFit)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

